Introduction to genomics and DNA methylation

### 2015-05-01 Rmarkdown/biology_background.Rmd
Update thesis structure.
Add brief captions

\chapter{Biology background}\label{chap:biology_background}

@@

The double helix is the most common, and most famous, structure of DNA. In the double helix, the two strands of DNA run in opposite directions to each other and are therefore _anti-parallel_. One strand is called the 5' strand (pronounced "5 prime strand" and also known as the sense strand, Crick strand or top strand) and the other strand called the 3' strand (pronounced "3 prime strand" and also known as the antisense strand, Watson strand or bottom strand). Along each strand of the double helix are the four DNA nucleobases (_bases_): adenine (_A_), cytosine (_C_), guanine (_G_) and thymine (_T_). These bases form _complementary_ base pairings, _A_ with _T_ and _C_ with _G_, along the DNA double helix. This is illustrated in Figure \ref{fig:double_helix}.

![Simple diagram of double-stranded DNA showing complementary base pairings. By Forluvoft (Own work) [Public domain], via Wikimedia Commons [http://commons.wikimedia.org/wiki/File%3ADNA_simple2.svg](http://commons.wikimedia.org/wiki/File%3ADNA_simple2.svg)\label{fig:double_helix}](../figures/DNA_simple2.pdf)

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.3\textwidth]{../figures/5-Methylcytosine.pdf}
\caption{Chemical structure of 5-methylcytosine. "5-methylcytosine". By Yikrazuul (Own work) [Public domain], via Wikimedia Commons \url{http://commons.wikimedia.org/wiki/File:5-Methylcytosine.svg}}
\label{fig:5mC}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/dinucleotide_frequency_hg19.pdf}
\caption{Observed:expected ratio of dinucleotides in the human reference genome (hg19). The expected frequency is computed under an 'independence' model, based on the observed frequencies of each base.}
\label{fig:dinucleotide_frequency_hg19}
\end{figure}

@@

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{../figures/IPD_ECDF_hg19.pdf}
\caption{Plot of the empirical cumulative distribution function of distance between adjacent CpGs in the human reference genome (hg19). The observed distances are contrasted with those under the 'expected' model whereby CpGs are distributed uniformly at random with probability equal to the observed frequency of Cpgs on each chromosome.}
\label{fig:IPD_ECDF_hg19}
\end{figure}

@@

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{../figures/IPD_frequency_polygon_hg19.pdf}
\caption{The frequency of the observed distances between adjacent CpGs in the human reference genome (hg19).}
\label{fig:IPD_frequency_polygon_hg19}
\end{figure}

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/bisulfite-conversion.pdf}
\caption{The effect of bisulfite-treatment of DNA. The double-stranded DNA is denatured and each strand undergoes bisulfite-treatment. Methylated cytosines remain as cytosines while unmethylated cytosines become uraciles. These bisulfite-converted DNA strands then undergo PCR amplification which converts the uracils to thymines. Note that while there are four possible PCR products, some bisulfite-sequencing protocols do not sample PCR products from the \textit{CTOT} or \textit{CTOB} strands. \textit{OT} = original top strand; \textit{OB} = original bottom strand; \textit{CTOT} = complementary to the original top strand; \textit{CTOB} = complementary to the original bottom strand. This figure is adapted from \citet{Krueger:2012ks}.}
\label{fig:bisulfite-conversion}
\end{figure}

@@

\begin{table}[h]
\centering
\begin{tabular}{@{}rc@{}}
\toprule
Sample              & Reads with no CpG \\ \midrule
ADS-iPSC            & $35 \%$                \\
ADS-adipose         & $37 \%$                \\
ADS                 & $39 \%$                \\
H9\_Laurent         & $39 \%$                \\
FF-iPSC\_19.11      & $48 \%$                \\
FF-iPSC\_19.7       & $49 \%$                \\
H1+BMP4             & $49 \%$                \\
FF-iPSC\_6.9        & $50 \%$                \\
IMR90-iPSC          & $50 \%$                \\
H9                  & $51 \%$                \\
FF-iPSC\_19.11+BMP4 & $52 \%$                \\
FF                  & $53 \%$                \\
IMR90\_r2           & $57 \%$                \\
IMR90\_r1           & $58 \%$                \\
H1\_r2              & $59 \%$                \\
H1\_r1              & $62 \%$                \\
HSF1                & $72 \%$                \\ \bottomrule
\end{tabular}
\caption{The percentage of reads that do not contain any CpGs for each sample in the \emph{Lister} dataset. Paired-end reads are counted as a single read and as such have a lower percentage of reads without CpGs. See Chapter \ref{chap:datasets} for a description of the \emph{Lister} dataset.}
\label{tab:reads_without_cpgs}
\end{table}

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/bisulfite_CT_SNP.pdf}
\caption{Sequence variation between the reference genome and the sample's genome can result in incorrect inference about the methylation state of the sample's genome. The locus in orange is a cytosine in the reference genome but a thymine in the sample's genome. Because the read is compared against the reference genome, it may incorrectly inferred to be an unmethylated cytosine. The locus in purple is a thymine in the reference genome but an unmethylated cytosine in the sample's genome. Because the read is compared against the reference genome it may incorrectly inferred to be a thymine.}
\label{fig:bisulfite_CT_SNP.pdf}
\end{figure}

>>>>

\chapter{Introduction to genomics and DNA methylation}\label{chap:biology_background}

@@

The double helix is the most common, and most famous, structure of DNA. In the double helix, the two strands of DNA run in opposite directions to each other and are therefore _anti-parallel_. One strand is called the 5' strand (pronounced "5 prime strand" and also known as the sense strand, Crick strand or top strand) and the other strand called the 3' strand (pronounced "3 prime strand" and also known as the antisense strand, Watson strand or bottom strand). Along each strand of the double helix are the four DNA nucleobases (_bases_): adenine (_A_), cytosine (_C_), guanine (_G_) and thymine (_T_). These bases form _complementary_ base pairings, _A_ with _T_ and _C_ with _G_, along the DNA double helix. This is illustrated in Figure \ref{fig:dna_simple2}.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.7\textwidth]{../figures/DNA_simple2.pdf}
\caption[DNA double helix]{Simple diagram of double-stranded DNA showing complementary base pairing. By Forluvoft (Own work) [Public domain], via Wikimedia Commons \url{http://commons.wikimedia.org/wiki/File\%3ADNA_simple2.svg}}
\label{fig:double_helix}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.3\textwidth]{../figures/5-Methylcytosine.pdf}
\caption[Chemical structure of 5-methylcytosine]{Chemical structure of 5-methylcytosine. "5-methylcytosine". By Yikrazuul (Own work) [Public domain], via Wikimedia Commons \url{http://commons.wikimedia.org/wiki/File:5-Methylcytosine.svg}}
\label{fig:5mC}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/dinucleotide_frequency_hg19.pdf}
\caption[Dinucleotide frequencies in the human genome]{Observed:expected ratio of dinucleotides in the human reference genome (hg19). The expected frequency is computed under an 'independence' model, based on the observed frequencies of each base.}
\label{fig:dinucleotide_frequency_hg19}
\end{figure}

@@

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{../figures/IPD_ECDF_hg19.pdf}
\caption[ECDF of distances between CpGs in hg19]{Plot of the empirical cumulative distribution function of distance between adjacent CpGs in the human reference genome (hg19). The observed distances are contrasted with those under the 'expected' model whereby CpGs are distributed uniformly at random with probability equal to the observed frequency of Cpgs on each chromosome.}
\label{fig:IPD_ECDF_hg19}
\end{figure}

@@

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{../figures/IPD_frequency_polygon_hg19.pdf}
\caption[Frequency of distances between CpGs in hg19]{The frequency of the observed distances between adjacent CpGs in the human reference genome (hg19).}
\label{fig:IPD_frequency_polygon_hg19}
\end{figure}

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/bisulfite-conversion.pdf}
\caption[Schematic of bisulfite-conversion]{The effect of bisulfite-treatment of DNA. The double-stranded DNA is denatured and each strand undergoes bisulfite-treatment. Methylated cytosines remain as cytosines while unmethylated cytosines become uraciles. These bisulfite-converted DNA strands then undergo PCR amplification which converts the uracils to thymines. Note that while there are four possible PCR products, some bisulfite-sequencing protocols do not sample PCR products from the \textit{CTOT} or \textit{CTOB} strands. \textit{OT} = original top strand; \textit{OB} = original bottom strand; \textit{CTOT} = complementary to the original top strand; \textit{CTOB} = complementary to the original bottom strand. This figure is adapted from \citet{Krueger:2012ks}.}
\label{fig:bisulfite-conversion}
\end{figure}

@@

\begin{table}[h]
\centering
\caption[Percentage of reads without CpGs]{The percentage of reads that do not contain any CpGs for each sample in the \emph{Lister} dataset. Paired-end reads are counted as a single read and as such have a lower percentage of reads without CpGs. See Chapter \ref{chap:datasets} for a description of the \emph{Lister} dataset.}
\label{tab:reads_without_cpgs}
\begin{tabular}{@{}rc@{}}
\toprule
Sample              & Reads with no CpG \\ \midrule
ADS-iPSC            & $35 \%$                \\
ADS-adipose         & $37 \%$                \\
ADS                 & $39 \%$                \\
H9\_Laurent         & $39 \%$                \\
FF-iPSC\_19.11      & $48 \%$                \\
FF-iPSC\_19.7       & $49 \%$                \\
H1+BMP4             & $49 \%$                \\
FF-iPSC\_6.9        & $50 \%$                \\
IMR90-iPSC          & $50 \%$                \\
H9                  & $51 \%$                \\
FF-iPSC\_19.11+BMP4 & $52 \%$                \\
FF                  & $53 \%$                \\
IMR90\_r2           & $57 \%$                \\
IMR90\_r1           & $58 \%$                \\
H1\_r2              & $59 \%$                \\
H1\_r1              & $62 \%$                \\
HSF1                & $72 \%$                \\ \bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/bisulfite_CT_SNP.pdf}
\caption[Sequence variation confounds methylation calling]{Sequence variation between the reference genome and the sample's genome can result in incorrect inference about the methylation state of the sample's genome. The locus in orange is a cytosine in the reference genome but a thymine in the sample's genome. Because the read is compared against the reference genome, it may incorrectly inferred to be an unmethylated cytosine. The locus in purple is a thymine in the reference genome but an unmethylated cytosine in the sample's genome. Because the read is compared against the reference genome it may incorrectly inferred to be a thymine.}
\label{fig:bisulfite_CT_SNP.pdf}
\end{figure}
